Tutorial_4

### 2024-12-11 Tutorial_4.Rmd
RR updated tutorial 4.

```{r load-pkg, results='hide'}
pkg <- c("cmdstanr", "standist", "tidyverse", "RColorBrewer", "patchwork", 
         "brms", "tidybayes", "bayesplot", "future", "parallel", "VGAM", "faux")

@@

@@

For Lisa's faux package, see here: [https://debruine.github.io/faux/](https://debruine.github.io/faux/).
The vignettes provided by the faux package are particularly helpful and worth following before you take a deep dive in the below code.

@@

1. Fit a regression model to an existing dataset.
2. Use the regression model parameters to simulate one new dataset.
3. Write a function to create 1000s of datasets and vary parameters of interest (e.g., sample size, trial count, effect size).
4. Summarise the simulated data to estimate likely power or precision of the research design options.

@@

The basic workflow above is a modified version of that recommended by Solomon and Lisa.
In a typical workflow, once the datasets are simulated, a regression model would be fit to each datatset.
Then, the parameters from each model would be summarised to make a judgment about the likely precision or power of the desing.
However, in the case of Model_4 in this tutorial, which took ~8 hours to build, we are not able to easily build 1000 models per variation in sample size and/or trial count, as that would take an unreasonably long time.
Instead, we choose to take a simpler approach and summarise the data per bin and condition, rather than the parameter estimates from the model.
Although this is not as satisfying, we feel it can still be a useful guide to set expectations. 
To further save time and computational resources, we also choose to use a much simpler model than Model_4, in order to outline the basic approach to planning via data simulation. 
The details on this simpler model structure are provided below.

@@

Here, we again use data form Panis and Schmidt (2016), but to make things simpler, we focus on only 6 timebins (timebins 4-9) and two conditions (con and inc).

@@

```{r load-data}
data <- read_csv("Tutorial_4_planning/data/tidyppp_Exp1_n6.csv") %>%
  ## no mask condition only, neutral prime removed, timebins 4-9 only
  filter(no_ma == 1, no_pr < 1, timebin %in% c(4:9)) %>% 
  ## remove unnecessary columns
  select(-c(bl, target_dir, corresp, resp, respac, rel_ma, irr_ma, ran_ma, no_ma)) %>%
  ## rename the error column
  rename(error = resperr) %>% 
  ## create a new variable and some factors
  mutate(prime = if_else(con_pr == 1, "con", "inc"),
         prime = factor(prime,
                        levels = c("con", "inc")),
         timebin = factor(timebin)) %>% 
  ## simplify the data by selecting a few variables only
  select(pid, timebin, prime, outcome)
head(data)
str(data)
```

@@

```{r calculate-summary-data}
data_pid <- data %>% 
  group_by(pid, timebin, prime) %>% 
  summarise(outcome = mean(outcome))
head(data_pid)

data_group <- data %>% 
  group_by(timebin, prime) %>% 
  summarise(outcome = mean(outcome))
head(data_group)
```

@@

So, the approach here is to build an initial model using an index coding approach, but with a much simpler dataset than has been used in prior tutorials.

Because we are building Bayesian regression models, which generate a posterior distribution per condition of interest in our index coding approach, we can simply calculate contrasts of interests and associated prevision or interval estimates (95% quantile intervals, for example) to address any questions we may have about comparisons between conditions, such as inc vs con in a particular timebin.

@@

For an index coding model, priors will be set differently to how they were set in the earlier tutorials, because the earlier tutorials did not use an index coding approach.
Using an index coding approach, each condition has a prior for the mean of that condition.
And as we know from plot 3.1 above, the condition averages across the six time bins selected range from near zero on the hazard probability scale to 0.50. 
And in principle, the more general case could involve any hazard value from zero to 1.
Therefore, if we try to stick to a weakly informative approach to priors [see here for more details](https://github.com/stan-dev/stan/wiki/prior-choice-recommendations), then priors for the cloglog values should cover a fairly broad range.

@@

@@

- model checking by looking at the convergence of chains and model diagnostics. 
We'll skip this step for now, just to move on to the main focus of planning and data simulation.

- Summarise and visualise the parameter estimates and/or posterior predictions.
Below, we quickly summarise fixed effects from the model in cloglog values, and we also convert them to hazard values, which tend to be far easier to interpret.

@@

# plot
p4.1 <- ggplot(tidy_fixed, aes(x = bin, y = value, fill=prime)) +  
  stat_halfeye(alpha=0.7) +
  labs(title = "Model bi cloglog",
       x = "timebin", y = "cloglog") +
  scale_fill_brewer(palette = "Dark2") 
p4.1

ggsave ("Tutorial_4_planning/figures/index_cloglog.jpeg",
        width = 6, height = 4)
```

@@

```{r plot-bi-hazards}
p4.2 <- ggplot(tidy_haz, aes(x = bin, y = hazard,
                                    fill=prime)) +  
  stat_halfeye(alpha=0.7) +
  labs(title = "hazard estimates",
       x = "timebin", y = "hazard") +
  scale_fill_brewer(palette = "Dark2") +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.25)) 
p4.2

ggsave ("Tutorial_4_planning/figures/index_hazard.jpeg",
        width = 6, height = 4)
```

@@

Note: For multi-level data, data simulation can be quite a convoluted process, and we therefore strongly recommend reading the power blogs by Solomon Kurz and working through the faux package vignettes, before getting stuck into the below code.

@@

## b1 to b6 = timebin 1 to 6.
## c = con
b1c = -4.66
b2c = -3.24
b3c = -2.32
b4c = -1.49
b5c = -1.21
b6c = -0.60

@@

```{r summary-d1-out}
d1_out_pid <- d1_out %>% 
  group_by(subj, condition, bin) %>% 
  summarise(outcome = mean(outcomen)) %>% 
  mutate(prime = if_else(condition == "cond1", "con", "inc"),
         prime = factor(prime))
d1_out_pid

d1_out_group <- d1_out %>% 
  group_by(condition, bin) %>% 
  summarise(outcome = mean(outcomen)) %>% 
  mutate(prime = if_else(condition == "cond1", "con", "inc"),
         prime = factor(prime))
d1_out_group
```

@@

this is the descriptive data per bin and condition for hazards
probability of a response/event, given that it has not happened yet

@@

Given that the effect of interest was ~80% reduction in Sven's 2016 paper, let's play around with a range of effect sizes from e.g., 0.25, 0.5, 0.75.
These would translate to a 75% reduction, 50% reduction and 25% reduction from con to inc.

@@

create a function to simulate data
These values are from the initial model and data from Sven's paper, but they can be modified when running the function, which we do below.

@@

vary trial count per condition and effect size, keep pid fixed at 10.
This will take some time, probably hours...go get a beer/coffee.
For the purposes of this tutorial, eval is set to false in the below chunks, to save time.

@@

```{r sim1-violin}
p9.1 <- ggplot(sx1_sim, 
                aes(x = bin, y = sim_mean_outcome, 
                    colour = condition)) +
  geom_jitter(data=sx1_exp, aes(y = exp_mean_outcome),
              alpha=0.5, width = 0.1, height = 0) +
  geom_line(aes(group = condition)) + 
  geom_point(size=2, colour = "black") +
  geom_errorbar(aes(ymin = sim_mean_outcome-sem*1.96, ymax = sim_mean_outcome+sem*1.96),
                width=.2, colour = "black") +
  scale_colour_brewer(palette = "Dark2") +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.25)) +
  labs(x="timebin",
       y="outcome") + 
  facet_grid(fct_rev(rep_n)~b6c)
p9.1

@@

```{r plot-ratio-values}
p9.2 <- ggplot(sx1_sim_check, 
                aes(x = b6c, y = sim_ratio)) +
  geom_point(size=1.5) +
  geom_errorbar(aes(ymin = sim_ratio-sim_sem*1.96, ymax = sim_ratio+sim_sem*1.96)) + 
  geom_hline(yintercept = c(0.25, 0.5, 0.75), colour = "red") +
  scale_colour_brewer(palette = "Dark2") +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.25)) +
  labs(x="timebin",
       y="outcome") + 
  facet_wrap(~rep_n)
p9.2
```

@@

basic idea - calculate the difference score in hazard between con and inc per bin and then report how many lower bound quantiles exclude zero? This of course is not a model-based estimate. But I think it should give a good guide to what we might expect.

@@

- effects of con vs inc that are as large as previously reported by Panis et al., 2016 would be easy to detect with 100 trials and 10 pids. And probably less data than that also.

@@

# 11. Run some more follow-up sims #

Let's try a 50%, 60% and 70% reduction across N=10,15 and 20.

@@

vary effect size and pid.
Again, as before, this will take some time, probably hours...go get a beer/coffee.
And again, eval is set to FALSE, to save time for the tutorial.

@@

```{r sim2-violin}
p11.1 <- ggplot(sx2_sim, 
                aes(x = bin, y = sim_mean_outcome, 
                    colour = condition)) +
  geom_jitter(data=sx2_exp, aes(y = exp_mean_outcome),
              alpha=0.5, width = 0.1, height = 0) +
  geom_line(aes(group = condition)) + 
  geom_point(size=2, colour = "black") +
  geom_errorbar(aes(ymin = sim_mean_outcome-sem*1.96, ymax = sim_mean_outcome+sem*1.96),
                width=.2, colour = "black") +
  scale_colour_brewer(palette = "Dark2") +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.25)) +
  labs(x="timebin",
       y="outcome") + 
  facet_grid(fct_rev(subj_n)~b6c)
p11.1

@@

```{r sim2-ratio-values}
p11.2 <- ggplot(sx2_sim_check, 
                aes(x = b6c, y = sim_ratio)) +
  geom_point(size=1.5) +
  geom_errorbar(aes(ymin = sim_ratio-sim_sem*1.96, ymax = sim_ratio+sim_sem*1.96)) + 
  geom_hline(yintercept = c(0.3, 0.4, 0.5), colour = "red") +
  scale_colour_brewer(palette = "Dark2") +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.25)) +
  labs(x="timebin",
       y="outcome") + 
  facet_wrap(~subj_n)
p11.2
```

@@

basic idea - calculate the difference score in hazard between con and inc per bin and then report how many lower bound quantiles exclude zero? This of course is not a model-based estimate. But I think it should give a good guide to what we might expect.

>>>>

```{r load-pkg, results='hide'}
pkg <- c("cmdstanr", "standist", "tidyverse", "RColorBrewer", "patchwork", 
         "brms", "tidybayes", "bayesplot", "future", "parallel", "VGAM", "faux",
         "rstan")

@@

rstan_options(auto_write = TRUE)

@@

For Lisa's faux package, see here: [https://debruine.github.io/faux/](https://debruine.github.io/faux/).
The vignettes provided by the faux package are particularly helpful and worth following before you take a deep dive in the below code. 
There is also an associated [journal article](https://doi.org/10.1177/2515245920965119), which is worth reading.

@@

1. Fit a regression model to existing data.
2. Use the regression model parameters to simulate new data.
3. Write a function to create 1000s of datasets and vary parameters of interest (e.g., sample size, trial count, effect size).
4. Summarise the simulated data to estimate likely power or precision of the various research designs.

@@

The basic workflow above is a modified version of that recommended by Solomon and Lisa.
In a typical workflow, once the datasets are simulated, a regression model would be fit to each dataset.
Then, the parameters from each model would be summarised to make a judgment about the likely precision or power of the design.
However, in the case of some of the models in this tutorial, which can take several hours to build, we are not able to easily build 1000 models per variation in parameter setting, such as sample size and/or trial count, as that would take an unreasonably long time on a desktop machine.
Instead, we choose to take a simpler approach and summarise the data per bin and condition, rather than the parameter estimates from the model.
Although this is not as satisfying, we feel it can still be a useful guide to set expectations. 
To further save time and computational resources, for the purposes of this planning tutorial, we also choose to use a smaller dataset than was used in the earlier tutorials. 
The details on this simpler dataset and modelling procedure are provided below.

@@

Here, we again use data form Panis and Schmidt (2016), but to make things simpler, we focus on only 6 time bins (time bins 4-9) and two prime conditions (congruent and incongruent).

@@

```{r load-data}
data <- read_csv("Tutorial_4_planning/data/tidyppp_Exp1_n6.csv") %>%
  ## no mask condition only, neutral prime removed, time bins 4-9 only
  filter(no_ma == 1, no_pr < 1, timebin %in% c(4:9)) %>% 
  ## remove unnecessary columns
  select(-c(bl, target_dir, corresp, resp, respac, rel_ma, irr_ma, ran_ma, no_ma)) %>%
  ## rename the error column
  rename(error = resperr) %>% 
  ## create a new variable and some factors
  mutate(prime = if_else(con_pr == 1, "con", "inc"),
         prime = factor(prime,
                        levels = c("con", "inc")),
         timebin = factor(timebin)) %>% 
  ## simplify the data by selecting a few variables only
  select(pid, timebin, prime, outcome)
head(data)
str(data)
```

@@

```{r calculate-summary-data}
data_pid <- data %>% 
  group_by(pid, timebin, prime) %>% 
  summarise(outcome = mean(outcome, na.rm = TRUE))
head(data_pid)

data_group <- data %>% 
  group_by(timebin, prime) %>% 
  summarise(outcome = mean(outcome, na.rm = TRUE))
head(data_group)
```

@@

The approach here is to build an initial model using an index coding approach, but with a subset of the data that was used in prior tutorials.

Because we are building Bayesian regression models, which generate a posterior distribution per condition of interest in our index coding approach, we can simply calculate contrasts of interests and associated precision or interval estimates (95% quantile intervals, for example) to address any questions we may have about comparisons between conditions, such as incongruent vs congruent in a particular time bin.

@@

Using an index coding approach, each condition has a prior for the mean of that condition.
And as we know from plot 3.1 above, the condition averages across the six time bins selected range from near zero on the hazard probability scale to 0.50. 
And in principle, the more general case could involve any hazard value from zero to 1.
Therefore, if we try to stick to a weakly informative approach to priors [see here for more details](https://github.com/stan-dev/stan/wiki/prior-choice-recommendations), then priors for the cloglog values should cover a fairly broad range.

@@

For simplicity, we use these normal(0,2) priors for the mean per condition.
However, if we were doing this for real, we would probably choose different priors that are similar to ones that were discussed in Tutorial_2a as well as in Supplmentary Materials.

@@

- check the model by looking at the convergence of chains and other model diagnostics. 
We'll skip this step for now, just to move on to the main focus of planning and data simulation.

- Summarise and visualise the parameter estimates and/or posterior predictions.

Below, we quickly summarise fixed effects from the model in cloglog values, and we also convert them to hazard values, which tend to be far easier to interpret.

@@

# plot
p4.1 <- ggplot(tidy_fixed, aes(x = bin, y = value, fill=prime)) +  
  stat_halfeye(alpha=0.7) +
  labs(title = "Model bi cloglog",
       x = "time bin", y = "cloglog") +
  scale_fill_brewer(palette = "Dark2") 
p4.1

ggsave ("Tutorial_4_planning/figures/index_cloglog.jpeg",
        width = 6, height = 4, dpi = 800)
```

@@

```{r plot-bi-hazards}
p4.2 <- ggplot(tidy_haz, aes(x = bin, y = hazard,
                                    fill=prime)) +  
  stat_halfeye(alpha=0.7) +
  labs(title = "hazard estimates",
       x = "time bin", y = "hazard") +
  scale_fill_brewer(palette = "Dark2") +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.25)) 
p4.2

ggsave ("Tutorial_4_planning/figures/index_hazard.jpeg",
        width = 6, height = 4, dpi = 800)
```

@@

Note: For multi-level data, data simulation can be quite a convoluted process, and we therefore strongly recommend reading the power blogs by Solomon Kurz and working through the faux package [vignettes](https://debruine.github.io/faux/index.html), before getting stuck into the below code.

@@

## b1 to b6 = time bin 1 to 6.
## c = con
b1c = -4.66
b2c = -3.24
b3c = -2.32
b4c = -1.49
b5c = -1.21
b6c = -0.60

@@

```{r summary-d1-out}
d1_out_pid <- d1_out %>% 
  group_by(subj, condition, bin) %>% 
  summarise(outcome = mean(outcomen, na.rm = TRUE)) %>% 
  mutate(prime = if_else(condition == "cond1", "con", "inc"),
         prime = factor(prime))
d1_out_pid

d1_out_group <- d1_out %>% 
  group_by(condition, bin) %>% 
  summarise(outcome = mean(outcomen, na.rm = TRUE)) %>% 
  mutate(prime = if_else(condition == "cond1", "con", "inc"),
         prime = factor(prime))
d1_out_group
```

@@

this is the descriptive data per bin and condition for the hazard - probability of a response/event, given that it has not happened yet

@@

Given that the effect of interest was ~80% reduction in Sven's 2016 paper (Panis & Schmidt, 2016), let's play around with a range of effect sizes from e.g., 0.25, 0.5, 0.75.
These would translate to a 75% reduction, 50% reduction and 25% reduction from con to inc.

@@

create a function to simulate data
These values are from the initial model and data from Sven's paper, which we created above in section 4, but they can be modified when running the function, which we do below.

@@

vary trial count per condition and effect size, keep pid fixed at N=10.
This will take some time, probably hours...go get a beer/coffee.
For the purposes of this tutorial, eval is set to false in the below chunks, to save time.

@@

```{r sim1-violin}
p9.1 <- ggplot(sx1_sim, 
                aes(x = bin, y = sim_mean_outcome, 
                    colour = condition)) +
  geom_jitter(data=sx1_exp, aes(y = exp_mean_outcome),
              alpha=0.5, width = 0.1, height = 0) +
  geom_line(aes(group = condition)) + 
  geom_point(size=2, colour = "black") +
  geom_errorbar(aes(ymin = sim_mean_outcome-sem*1.96, ymax = sim_mean_outcome+sem*1.96),
                width=.2, colour = "black") +
  scale_colour_brewer(palette = "Dark2") +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.25)) +
  labs(x="time bin",
       y="outcome") + 
  facet_grid(fct_rev(rep_n)~b6c)
p9.1

@@

```{r plot-ratio-values}
p9.2 <- ggplot(sx1_sim_check, 
                aes(x = b6c, y = sim_ratio)) +
  geom_point(size=1.5) +
  geom_errorbar(aes(ymin = sim_ratio-sim_sem*1.96, ymax = sim_ratio+sim_sem*1.96)) + 
  geom_hline(yintercept = c(0.25, 0.5, 0.75), colour = "red") +
  scale_colour_brewer(palette = "Dark2") +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.25)) +
  labs(x="time bin",
       y="outcome") + 
  facet_wrap(~rep_n)
p9.2
```

@@

basic idea - calculate the difference score in hazard between con and inc per bin and then report how many lower bound 95% quantile intervals exclude zero? This of course is not a model-based estimate. But we think it should give a reasonable guide to what we might expect.

@@

- effects of con vs inc that are as large as previously reported by Panis & Schimdt (2016) would be easy to detect with 100 trials and 10 pids. And probably less data than that also.

@@

# 11. Run a follow-up simulation #

Let's try a 50%, 60% and 70% reduction across N=10, 15 and 20.

@@

vary effect size and pid.
Again, as before, this will take some time, probably hours...go get another beer/coffee.
And again, eval is set to FALSE, to save time for the tutorial.

@@

```{r sim2-violin}
p11.1 <- ggplot(sx2_sim, 
                aes(x = bin, y = sim_mean_outcome, 
                    colour = condition)) +
  geom_jitter(data=sx2_exp, aes(y = exp_mean_outcome),
              alpha=0.5, width = 0.1, height = 0) +
  geom_line(aes(group = condition)) + 
  geom_point(size=2, colour = "black") +
  geom_errorbar(aes(ymin = sim_mean_outcome-sem*1.96, ymax = sim_mean_outcome+sem*1.96),
                width=.2, colour = "black") +
  scale_colour_brewer(palette = "Dark2") +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.25)) +
  labs(x="time bin",
       y="outcome") + 
  facet_grid(fct_rev(subj_n)~b6c)
p11.1

@@

```{r sim2-ratio-values}
p11.2 <- ggplot(sx2_sim_check, 
                aes(x = b6c, y = sim_ratio)) +
  geom_point(size=1.5) +
  geom_errorbar(aes(ymin = sim_ratio-sim_sem*1.96, ymax = sim_ratio+sim_sem*1.96)) + 
  geom_hline(yintercept = c(0.3, 0.4, 0.5), colour = "red") +
  scale_colour_brewer(palette = "Dark2") +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.25)) +
  labs(x="time bin",
       y="outcome") + 
  facet_wrap(~subj_n)
p11.2
```

@@

Same basic idea as in Simulation 1.
